---
title: "EDA"
author: "Tuan Pham"
date: "2023-04-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1. Load the Dataset

```{r}
# Load packages
library(tidyverse)
library(ggplot2)
library("scales")

```
```{r}
# Load a dataset
data <- read_csv('C:/Users/arsen/OneDrive/Desktop/ames.csv')
head(data)
```
```{r}
#View a structure of the data
str(data)
```


2. Explore Data Distributions

Sale Price
```{r}
#Set up a plot
ggplot(data, aes(x=SalePrice)) +
  geom_histogram()+
  scale_x_continuous(labels = comma)+
  geom_vline(xintercept=mean(data$SalePrice), color="black") +
  labs(title="Distribution of Sale Prices", x="Sale Price", y="Number of Houses")
```
```{r}
#Descriptive statistics of SalePrice
summary(data$SalePrice)
```
Looks like a log normal distribution. Most houses in this sample are
clustered around the median value of $163,000, but the higher-end
homes are pulling the mean up to over $180,000

Total Rooms Above Grade
```{r}
#Set up a plot
ggplot(data,aes(x=TotRmsAbvGrd)) +
  geom_histogram()+
   geom_vline(xintercept=mean(data$TotRmsAbvGrd), color="black") +
  labs(title="Distribution of Total Rooms Above Grade", x="Total Rooms (Does Not Include Bathrooms)", y="Number of Houses")
```
```{r}
#Descriptive Statistics of TotRmsAbvGrd
summary(data$TotRmsAbvGrd)
```
The number of rooms in houses is approximately normally distributed, 
with a mean and median around 6 rooms. There are some houses with
twice as many rooms as the average, but overall the distribution is
less skewed than the sale price distribution

Overall Condition
```{r}
#Set up a plot
ggplot(data,aes(x=OverallCond)) +
  geom_histogram()+
   geom_vline(xintercept=mean(data$OverallCond), color="black") +
  labs(title="Distribution of Overall Condition of Houses on a 1-10 Scale", x="Condition of House", y="Number of Houses")
```
```{r}
#Descriptive Statistics of OverallCond
summary(data$OverallCond)
```
Most homes have a condition of 5. It seems like we should
treat this as a categorical rather than numeric variable,
since the difference between conditions is so abrupt

3. Explore Differences between Subsets

```{r}
#Create subsets based on specific conditions
below_average_condition <- data %>% filter(OverallCond < 5)
average_condition <- data %>% filter(OverallCond == 5)
above_average_condition <- data %>% filter(OverallCond > 5)
```

```{r}
#Set up a plot
ggplot() +
  geom_histogram(data=above_average_condition, aes(x=SalePrice, fill="above average condition"), alpha=0.5) +
  geom_histogram(data=average_condition, aes(x=SalePrice, fill="average condition"), alpha=0.3) +
  geom_histogram(data=below_average_condition, aes(x=SalePrice, fill="below average condition"), alpha=0.5) +
  scale_x_continuous(labels = comma) +
  labs(title="Distributions of Sale Price Grouped by Condition", x="Sale Price", y="Number of Houses", fill="") +
  guides(fill = guide_legend(reverse=TRUE)) +
  scale_fill_manual(values=c("cyan", "gray", "yellow"), labels=c("above average condition", "average condition", "below average condition"))

```
First, we note again that the majority of the houses have average
condition, then about 1/3 have above average condition, then less
than 10% have below average condition.

As we might expect, the average condition therefore contains houses
across a broader spectrum of the sale price range than either the
below-average or above-average houses.

Another unsurprising finding is that below-average condition houses
have a price distribution that is much lower than average or
above-average condition houses.

But what might be surprising is that above-average condition houses
do not seem to have higher average sale prices than average condition
houses. In fact, above-average condition houses seem more clustered
around a particular price range, especially the $100,000 to $200,000
range, whereas average condition houses are more frequent above
$200,000. We might want to investigate further to understand what
kinds of houses are rated as above-average condition, since this goes
against a standard assumption that better condition would mean higher
cost.

4. Explore Correlations

The most positively correlated with SalePrice
```{r}
#Extract numeric columns from the dataset
numeric_cols <- sapply(data, is.numeric)
numeric_cols
data_numeric <- data[, numeric_cols]
data_numeric

# Get a list of correlations with SalePrice, sorted from smallest to largest
correlation_series <- sort(cor(data_numeric)[,'SalePrice'])

# Select second to last correlation, since the highest (last)
# correlation will be SalePrice correlating 100% with itself
max_corr_value <- correlation_series[length(correlation_series) - 1]
max_corr_column <- names(correlation_series)[length(correlation_series) - 1]

# Print the most positively correlated column and its maximum correlation value
cat("Most Positively Correlated Column: ", max_corr_column, "\n")
cat("Maximum Correlation Value: ", max_corr_value, "\n")
```

The most negatively correlated with SalePrice
```{r}
# Print the most negatively correlated column and its minimum correlation value
min_corr_value <- correlation_series[1]
min_corr_column <- names(correlation_series)[1]

cat("Most Negatively Correlated Column: ", min_corr_column, "\n")
cat("Minimum Correlation Value: ", min_corr_value, "\n")
```

Once we have enough information, we creates box-plots of the relevant columns
```{r}
# Plot distribution of column with highest correlation
ggplot(data, aes(x = reorder(as.character(OverallQual),SalePrice), y = SalePrice)) +
  geom_boxplot() +
  scale_y_continuous(labels = comma) +
  labs(title = "Overall Quality vs. Sale Price",
       x = "Overall Quality",
       y = "Sale Price")

# Plot distribution of column with most negative correlation
ggplot(data, aes(x = reorder(as.character(KitchenAbvGr),SalePrice), y = SalePrice)) +
  geom_boxplot() +
  scale_y_continuous(labels = comma) +
  labs(title = "Number of Kitchens vs. Sale Price",
       x = "Number of Kitchens Above Ground",
       y = "Sale Price")
```

5. Engineer and Explore a New Feature

```{r}
table(data$YrSold)
```
```{r}
#Create a new column
df <- data %>%
  mutate(Age = YrSold - YearBuilt)

#Set up a plot
ggplot(df, aes(x = Age, y = SalePrice)) +
  geom_point(alpha = 0.3, color = "green") +
  labs(title = "Home Age vs. Sale Price", 
       x = "Age of Home at Time of Sale", 
       y = "Sale Price") +
  theme_bw()
```
In general, newer houses appear to be more valuable, with value increasing
as homes age. Interestingly the variance seems to increase once the home
age goes over 100 years, with several above-average sale prices and fewer
home sales in general.

We are also seeing potential housing booms and busts over the past decades,
indicated by e.g. relatively few 20-year-old houses compared to 25-year-old houses
being sold. We might find something interesting if we investigae this further.
